Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-07 ms.tex
full integration of modifier model; some small revisions

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.

Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
Our result shows that if below-average individuals produce more variation than above-average individuals, then the population mean fitness increases.

@@

The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring), independent of the frequencies of the other types, and is fully described by the matrix $M$. Specifically, transmission infidelity causes type transitions.
Therefore, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

@@

We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine the rate of transition out of the different types ($C_k$).
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.

% Modifier model
\paragraph*{Modifier model.}
Considering the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
the rest of the model parameters are the same as in Eq.~\ref{eq:model};
and the frequencies in the next generation, $f'$ and $g'$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.

@@

% Mean fitness principle
Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v
\end{equation}

The following theorem presents a \emph{mean fitness principle} for the
sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
in $C_k$ the probability of transition from $A_k$.

\begin{lemma}[Mean fitness principle]
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:sign_theorem}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Then,
\begin{equation}\label{eq:theorem_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k D_k + v_k D_k u\tr[M]_k = \\
v_k D_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{eq:theorem_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - D_k),
\end{equation}
which completes the proof.
\end{proof}

The above theorem provides a condition for the effect of a modifier in control of $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.

@@

An interesting intermediate result in Eq.~\ref{eq:theorem_intermediate} is

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.

Given $0 < D_k \le 1$, from Eq.~\ref{eq:theorem_intermediate} we can get the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of the theorem,
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

@@

We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{eq:sign_theorem_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:sign_theorem}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}), and $D_i \le 1$ for all $i$.

@@

We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
Could such a modifier increase in frequency when initially rare in the population?
To answer this question we analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of modifier allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
\end{equation}
The zero block matrices are due to complete linkage between the modifier and the types $A_k$.

@@

\begin{theorem}[Evolution of increased genetic variation]
The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $\cl_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}
In addition, the leading eigenvalue is $\lambda=1$ when transition rates are equal $\tilde{C}_k=C_k$ for all $k$.

Therefore, an initially rate modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

@@

Now, applying Eq.~\ref{eq:sign_theorem} to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
\end{equation}
and evaluating this expression at $\lambda=1$ (the case for $\tilde{C}_k = C_k$) gives us
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
sign(\bar{w}^* - D_k),
\end{equation}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-fitness individuals increase the rate at which variation is generated.
We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our \emph{mean fitness principle} will hold.

Eq.~\ref{eq:theorem} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{eq:sign_theorem} and~\ref{eq:sign_theorem_threshold}.

@@

General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
Here we have provided another general example, which suggests that if maladapted or below-average individuals increase the rate at which variation is generated, the population mean fitness will increase.

>>>>

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Lessard1990}.

Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of the individual, and specifically on its fitness.
Our results show that the population mean fitness increases if below-average individuals produce more variation than above-average individuals,
and that modifier alleles that induce below-average to produce more variation are favored by natural selection.

@@

The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

@@

We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$ the rates of transition out of the different types.
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

% Modifier model
\paragraph*{Modifier model.}
Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for the absence of allele $M$, with $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.
For simplicity of the presentation we focus on one modifier allele in these examples.

@@

% Mean fitness principle
Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}

The following lemma presents a \emph{mean fitness principle} for the
sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
in $C_k$, the probability of transition from $A_k$.

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k D_k + v_k D_k u\tr[M]_k = \\
v_k D_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - D_k),
\end{equation}
which completes the proof.
\end{proof}

The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.

@@

An interesting intermediate result in Eq.~\ref{mfp_intermediate} is

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.

Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we can get the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

@@

We turn our attention to a modifier allele that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.

@@

We determined a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
Could such an allele increase in frequency when initially rare in the population?
To answer this question we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
\end{equation}
The zero block matrices are due to complete linkage between the modifier and the types $A_k$.

@@

\begin{theorem}[Evolution of increased genetic variation]
The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$, when evaluated at $\tilde{C}_k=C_k$, is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}
In addition, the leading eigenvalue is $\lambda=1$ when transition rates of modifier alleles $m$ and $M$ are equal ($\tilde{C}_k=C_k$ for all $k$).

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

@@

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
\end{equation}
and evaluating this expression for $\tilde{C}_k = C_k$), i.e. at $\lambda=1$, gives us
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
sign(\bar{w}^* - D_k),
\end{equation}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-average individuals increase the rate at which variation is generated. Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.

We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.

Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.

@@

General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
Here we have provided another general example, which suggests that a modifier allele that prompts maladapted or below-average individuals increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
